\textls[-15]{\thetitle \mbox{(Technical Report)}}

### 2022-06-14 canonicitymodel.tex
Reconsiderate phrase of "Extensible"
Save.

We follow \citep{sterling2019algebraic,coquand2018canonicity} to construct the canonicity model. We will first specify the formulation of the basic canonicity theorem.

We will resort to both conventional math notation and Agda syntax to explain the main idea of canonicity model.

% After gluing, we figure out the following "canonicity" theorem as well
% the canonical form of extensible inductive type, linkage and boolean

>>>>

We follow the reducibility argument from \citep{coquand2018canonicity,sterling2019algebraic} to construct the canonicity model. We will first specify the formulation of the most basic canonicity theorem.

\begin{theorem}[Canonicity]
  For a closed boolean term $\goodTerm{\cdot}{t}{\mathbb{B}}$, we have $\goodTerm{\cdot}{t \equiv tt}{\mathbb{B}}$ or $\goodTerm{\cdot}{t \equiv ff}{\mathbb{B}}$ hold.
\end{theorem}

Canonicity is one of the basic criteria to consider a dependent type theory to be ``a programming language or as a computational foundation for mathematics''\footnote{See nlab explanation: \href{https://ncatlab.org/nlab/show/canonical+form}{Canonical Form}}.
% cite https://ncatlab.org/nlab/show/canonical+form here
We can even argue that, if this theorem is proven in a constructive meta-logic\footnote{QIIT does have computational content}, 
% citation needed
then by Curry-Howard Correspondence, this theorem provides a big-step interpreter for closed term of boolean type.

Now we sketch out the big picture of the proof. The canonicity model $"Tm"^C~\Gamma^C~"T"^C$ will map each syntax piece $("t : Tm Γ T") \mapsto "(t, tₚ) :" "Tm"^C~\Gamma^C~"T"^C = \sum~("t : Tm Γ T")."Tm"_p~\Gamma^C~"T"^C~"t"$ to a dependent pair\footnote{This $"Tm"_p$ is usually called dependent model.}, of which the first part will be the same as the input syntax. Using this mapping on closed boolean term $"t : Tm ⋅ "\mathbb{B}$ will give us $"Tm"_p~\cdot^C~\mathbb{B}^C~"t"$, which will unfolds to our final goal $"t" \equiv "tt" + "t" \equiv "ff"$. Roughly, this can read as ``every closed boolean term is reducible''.  Thus once we can ``propagate'' $(\cdot)^C$ into every syntactic piece, and construct this model, we are done. In our (fake) Agda formulation, we use $(\cdot)_2$ to replace the notation of $(\cdot)^C$ because we consider most parts of this model are actually a dependent pair.

Another helpful analogy of the canonicity model is that $"Tm"_p~\Gamma^C~"T"^C~"t"$ can be understood as $"t" \in \{$ Reducible Terms of type $T$ (and $T_p$) in context $\Gamma$ (and $\Gamma_p$) $\}$. The latter is more conventional and the former can be understood as type theoretic encoding of the latter\footnote{We usually use predicate to encode "belong to" relation in type theory}. From this perspective, this part of the canonicity model is quite similar to how conventional (proof-irrelevant) logical relation is carried out --
% cite https://arxiv.org/abs/1907.11133
using the terminology of the conventional formulation, $\Gamma_p$ will be the set of substitution into closed reducible terms, and we will prove $t[\gamma] \in T_p$, a set of closed reducible term of type $T$ when $\gamma \in \Gamma_p$.

The proof can be found in appendix.

With the logical relation and the help of eta rule, we figure out the following canonical form
\begin{theorem}[Canonical Forms].
  \begin{itemize}
    \item if $\goodTerm{\cdot}{t}{El~("W"~*)}$ with $\goodTerm{\cdot}{*}{\bW~\sigma}$ and $\goodWSig{\cdot}{\sigma}{n}$, then $\goodTerm{\cdot}{t \equiv "Wsup"~j~a~b}{El~("W"~*)}$ for some $\goodTerm{\cdot}{a}{A}$ and $\goodTerm{\cdot, B[(id, a)]}{b}{El~("W"~*)}$ and $j < n$
    \item if $\goodTerm{\cdot}{t}{\mathbb{B}}$ then $\goodTerm{\cdot}{t \equiv "tt"}{\mathbb{B}}$ or $\goodTerm{\cdot}{t \equiv "ff"}{\mathbb{B}}$ 
    \item if $\goodTerm{\cdot}{t}{\cL\sigma}$ with $\goodSig{\cdot}{\sigma}{n}$, then $\goodTerm{\cdot}{t \equiv \mu^+~o~t}{\cL\sigma}$

      $\quad$ for some $\goodTerm{\cdot}{o}{\cL(p_1\nu~\sigma)}$ and $\goodTerm{\cdot, \cC(p_1\nu'~\sigma)}{t}{p_2\nu~\sigma}$
    \item if $\goodTerm{\cdot}{t}{\Sigma A B}$ then $\goodTerm{\cdot}{t = (a, b)}{\Sigma A B}$ with $\goodTerm{\cdot}{a}{A}$ and $\goodTerm{\cdot}{b}{B[(id, a)]}$
    \footnote{We emphasize the last one because $\cCt t$ is a dependent pair}
  \end{itemize}
\end{theorem}

% the canonical form of inductive type, linkage and boolean

### 2022-06-14 standardmodel.tex
Save.

We still encourage the reader to read the (fake) Agda version because most of the parts are similar but with more type annotation so it is clearer about the mathematical object doing the interpretation. Here, we need to clarify some ambiguity in our conventional math formulation. For example, when we wrote ``$\denotesS{\goodType{\Gamma}{\cC \sigma}{}}$ defined inductively on $\denotesS{\Sigma}$'', the best way to understand is to consider we are defining a function $\denotesS{\goodType{\Gamma}{\cC \_}{}}$ as the interpretation for $\goodType{\Gamma}{\cC \_}{}$ that mapping $\denotesS{\sigma}$ to $\denotesS{\goodType{\Gamma}{\cC \sigma}{}}$ also defined inductively on the length of the signature. The case is similar for $\denotesS{\cL \_}$, $\denotesS{\cCt \_}$ and  $\denotesS{\Sigma \ \_ \ \_}$.

\textit{This is actually the formulation in Agda} -- QIIT syntax consider $\cC : "Sig"~\Gamma~n \to "Ty"~\Gamma$ as a constructor and our model needs to interpret this constructor to a function $\cC^M : "Sig"^M \Gamma^M~n \to "Ty"^M~\Gamma^M$ with each judgement type also interpreted. Basically we are simply constructing a function that mapping our QIIT data-type to some other data -- $"Con"$ to $"Con"^M$, "Ty : Con → Set" to "Ty$^M$ : Con$^M$ → Set" and etc. What's more, since judgemental equality in our meta-theory is represented by equality in QIIT, and the well-defined-ness of this mapping-out function implies that ``judgementally equal syntax is interpreted by the same semantic in the model''.

@@

We will take some examples on the verification of validity of this model: 1. to verify the well-defined-ness of this model (i.e. judgemental equality is modeled by equality inside the model); 2. to verify compatibility lemma.

@@

>>>>

We encourage the reader to read the (fake) Agda-style proof in our appendix because most of the parts are similar but with more type annotation so it is clearer about the mathematical object doing the interpretation. Here, we need to clarify some ambiguity in our conventional formulation. For example, when we wrote ``$\denotesS{\goodType{\Gamma}{\cC \sigma}{}}$ defined inductively on $\denotesS{\Sigma}$'', the best way to understand is to consider we are defining a function $\denotesS{\goodType{\Gamma}{\cC \_}{}}$ as the interpretation for $\goodType{\Gamma}{\cC \_}{}$ that mapping $\denotesS{\sigma}$ to $\denotesS{\goodType{\Gamma}{\cC \sigma}{}}$ also defined inductively on the length of the signature. The case is similar for $\denotesS{\cL \_}$, $\denotesS{\cCt \_}$ and  $\denotesS{\Sigma \ \_ \ \_}$. \textbf{This is actually the formulation in Agda} -- QIIT syntax consider $\cC : "Sig"~\Gamma~n \to "Ty"~\Gamma$ as a constructor and our model needs to interpret this constructor to a function $\cC^S : "Sig"^S \Gamma^S~n \to "Ty"^S~\Gamma^S$ with each judgement type also interpreted.

Basically in Agda, we are simply constructing a function that mapping our QIIT data-type to some other data -- it maps $"Con"$ to $"Con"^S$, "Ty : Con → Set" to "Ty$^S$ : Con$^S$ → Set" and etc. What's more, since judgemental equality in our meta-theory is represented by equality in QIIT, and the well-defined-ness of this mapping-out function implies that ``judgementally equal syntax is interpreted by the same semantic in the model''.

@@

The verification of validity of this model is of course required but largely omitted and we refer the reader to the appendix. For example, to verify the well-defined-ness of this model (i.e. judgemental equality is modeled by equality inside the model). If we want to verify compatibility lemma.

@@

The full model formulated in (fake) Agda syntax can be found in the appendix.
